Automating Free Logic in HOL, with an Application in Category Theory\thanks{Mention DFG grants ...}

### 2017-03-23 FreeLogic/2017-Journal/article.tex
update

%%%%%%%%%%%%%%%%%%%%%%% file template.tex %%%%%%%%%%%%%%%%%%%%%%%%%
%
% This is a general template file for the LaTeX package SVJour3
% for Springer journals.          Springer Heidelberg 2010/09/16
%
% Copy it to a new file with a new name and use it as the basis
% for your article. Delete % signs as needed.
%
% This template includes a few options for different layouts and
% content for various journals. Please consult a previous issue of
% your journal as needed.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% First comes an example EPS file -- just ignore it and
% proceed on the \documentclass line
% your LaTeX will extract the file if required
\begin{filecontents*}{example.eps}
%!PS-Adobe-3.0 EPSF-3.0
%%BoundingBox: 19 19 221 221
%%CreationDate: Mon Sep 29 1997
%%Creator: programmed by hand (JK)
%%EndComments
gsave
newpath
  20 20 moveto
  20 220 lineto
  220 220 lineto
  220 20 lineto
closepath
2 setlinewidth
gsave
  .4 setgray fill
grestore
stroke
grestore
\end{filecontents*}
%
\RequirePackage{fix-cm}
%
%\documentclass{svjour3}                     % onecolumn (standard format)
%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
\documentclass[smallextended]{svjour3}       % onecolumn (second format)
%\documentclass[twocolumn]{svjour3}          % twocolumn
%
\smartqed  % flush right qed marks, e.g. at end of proof
%
\usepackage{amsmath}
\usepackage{xspace,stmaryrd,latexsym}
\usepackage{graphicx,tikz,rotating}
%
% \usepackage{mathptmx}      % use Times fonts if available on your TeX system
%
% insert here the call for the packages your document requires
%\usepackage{latexsym}
% etc.
%
% please place your own definitions here and don't use \def but
% \newcommand{}{}
%
% Insert the name of "your journal" with
% \journalname{myjournal}
%

\def\D{\mathbf{D}}
\def\E{\mathbf{E}}

@@

\newcommand{\imp}{\supset}
\newcommand{\biimp}{\equiv}
\newcommand{\allq}{\forall}
\newcommand{\exq}{\exists}
\newcommand{\seq}{\vdash}
\newcommand{\nec}{\Box} % necessarily
\newcommand{\pos}{\Diamond} % possibly
\newcommand{\ess}[2]{#1 \ \mathit{ess.} \ #2}
\newcommand{\NE}{\mathit{NE}}

@@

\title{Exploring Axiom Systems for Category Theory%\thanks{Grants or other notes
%about the article that should go on the front page should be
%placed here. General acknowledgments should be placed at the end of the article.}
}
\subtitle{Utilizing a 
 Novel Approach to Automate Free Logic in HOL}

@@

The purpose of this article is not claim any novel results in category
theory, but to demonstrate an elegant way to ``implement'' and utilise
reasoning in free logic, and to present respective experiments.

@@

To demonstrate the practical relevance of our new system, we present,
in the second part of this article, a stepwise development of axiom
systems for category theory by generalizing the standard axioms for a
monoid to a partial composition operation. Our purpose is not to make
or claim any contribution to category theory but rather to show how
formalizations involving the kind of logic required (free logic) can
be validated within modern proof assistants.

A total of eight different axiom systems is studied. The systems I-VI
are shown to be equivalent. The axiom system VII slightly modifies
axiom system VI to obtain (modulo notational transformation) the set
of axioms as proposed by Freyd and Scedrov in their textbook
``Categories, Allegories'' \cite{FreydScedrov90}, published in 1990;
see also Subsection \ref{subsec:FreydNotation} where we present their
original system.  While the axiom systems I-VI are shown to be
consistent, a constricted inconsistency result is obtained for system
VII (when encoded in free logic where free variables range over all
objects): We can prove $\exists x. \neg(E x) \Impl False$, where
$E$ is the existence predicate. Read this as: If there are undefined
objects, e.g. the value of an undefined composition $x\cdot y$
then we have falsity.  By contraposition, all objects (and thus all
compositions) must exist. But when we assume the latter, then the
axiom system VII essentially reduces categories to monoids.  We note
that axiom system V, which avoids this problem, corresponds to a set
of axioms proposed by Scott @{cite "Scott79"} in the 1970s. The
problem can also be avoided by restricting the variables in axiom
system VII to range only over existing objects and by postulating
strictness conditions.  This gives us axiom system VIII.

Our exploration has been significantly supported by series of experiments in which automated reasoning tools 
have been called from within the proof assistant Isabelle/HOL @{cite "Isabelle"} via the Sledgehammer 
tool @{cite "Sledgehammer"}. Moreover, we have obtained very useful feedback at various stages 
from the model finder Nitpick @{cite "Nitpick"} saving us from making several mistakes.

@@

\section{Preliminaries}
\label{sec:preliminaries}
\subsection{Free Logic}
Free logic \cite{Lambert60,Scottt} refers to a class of logic
formalisms that are free of basic existence assumptions regarding the
denotation of terms. Remember that terms in e.g. traditional classical
and intuitionistic predicate logics always denote an (existing) object
in a given (non-empty) domain $\D$, and that $\D$ is also exactly the set the
quantifiers range over. In free logic these
basic assumption are abolished. Terms do still denote objects in a
(non-empty) domain $\D$, but a (possibly empty) set $\E\subseteq \D$ is
chosen to characterize the subdomain of ``existing'' resp. ``defined''
objects in $\D$. Quantification is
now restricted to set $\E$ of existing/defined objects only. It is
obvious how this can be used to model undefideness and partiality:
problematic terms, e.g. division by zero or improper definite
descriptions, still denote, but they refer to undefined objects, that
is, objects $d$ in $\D\setminus \E$ lying outside of the scope of
quantification.

The particular notion of free logic as exploited in the remainder of
this article has been introduced by Scott \cite{Scott67}.  A graphical
illustration of this notion of free logic is presented in
Fig.~\ref{fig1}. It
employs a distinguished undefined object $\star$.  
% Free logic (and inclusive logic) has been proposed as an alternative
% to remedy these shortcomings. It distinguishes between a raw domain of
% possibly non-existing objects ‹❙D› and a particular subdomain ‹❙E› of ‹❙D›,
% containing only the ``existing'' entities. Free variables range over ‹❙D›
% and quantified variables only over ‹❙E›. Each term denotes in ‹❙D› but not
% necessarily in ‹❙E›. The particular notion of free logic as exploited below has been
% introduced by Scott @{cite "scott67:_exist"}. A graphical 
% illustration of this notion of free logic is presented in Fig.~\ref{fig1}.
\begin{figure}
\centering
\newcommand\firstellipse{(2,-5) ellipse (6cm and 4cm)}
\newcommand\secondellipse{(0,-5.3) ellipse (3.5cm and 2.5cm)}1
\begin{tikzpicture}
  \begin{scope}[fill opacity=1]
    \fill[gray!40] \firstellipse;
    \fill[gray!10] \secondellipse;
  \end{scope}
  \begin{scope}[very thick,font=\large]
    \draw \firstellipse node[font=\normalsize\bfseries] at (-.5,-5) {\underline{\textbf{E}: existing objects}};
    \draw \firstellipse node[font=\small] at (-.5,-5.5) {values of bound variables};
    \draw \secondellipse node[font=\normalsize\bfseries] at (4.3,-2.5) {\underline{\textbf{D}: raw objects}};
    \draw \secondellipse node[font=\small] at (4.3,-3) {values of free variables};
  \end{scope}
  \node[font=\normalsize\bfseries] at (6,-5) {$\star$};
  \node[font=\normalsize\bfseries] at (6,-5.3) {undefined};
\end{tikzpicture}
\caption{Illustration of the Semantical Domains of Free Logic \label{fig1}}
\end{figure}

@@

  The terms and formulas of $\FFOL$ are formally defined as the
  smallest sets such that:
\begin{enumerate}
\item each variable $x\in\V$ is a term of $\FFOL$, 
\item given any $n$-ary ($n\geq 0$) function symbol $f\in\F$ and terms
  $t_1,\ldots,t_n$ of $\FFOL$, then $f(t_1,\ldots,t_n)$ is a term of $\FFOL$, 
\item given terms $t_1$ and $t_2$ of $\FFOL$, then $t_1 = t_2$ is an
  (atomic) formula of $\FFOL$,
\item given any $n$-ary ($n\geq 0$) predicate symbol $p\in\P$ and terms
$t_1,\ldots,t_n$ of $\FFOL$, then $p(t_1,\ldots,t_n)$ is an (atomic)
formula of $\FFOL$, 
\item given formulas $r$ and $s$ of $\FFOL$, then $\neg r$, $r
  \Impl s$ and $\forall x. r$ are (compound) formulas of $\FFOL$
\item given a formula $r$ of $\FFOL$, then $\Descr x. r$ is a
  term of $\FFOL$ (definite  description)
\end{enumerate}
\end{definition}

Further formulas of $\FFOL$  can be defined as usual.

Regarding the semantics different options may be considered. For
example, instead of a possible empty set of existing objects $E$, we
could postulate non-emptiness of $E$. We e.g. also have a choice when
interpreting equations $t = t$ for undefined terms $t$; such equations
could be mapped to $\True$ or $\False$. Different combinations of such
choice points have been studied in the literature. Here we closely
follow the logic of Scott \cite{Scott67}.

\begin{definition}[Dual-domain (positive) semantics of $\FFOL$]
  A model structure for $\FFOL$ consists of a triple
  $\langle D, E, I, \star\rangle$, where $D$ is a non-empty raw domain of
  objects, $E\subseteq D$ a possible empty set of ``existing''
  resp. ``defined'' objects, and $I$ an interpretation function
  mapping 0-ary function symbols (constants) to defined objects $d\in E$, 
 0-ary predicate symbols (propositions) to $\True$ or $\False$, 
  $n$-ary function symbols (for $n\geq 1$) to $n$-ary functions
  $D \times \ldots \times D \longrightarrow D$ and $n$-ary predicate
  symbols (for $n\geq 1$) to $n$-ary relations
  $D \times \ldots \times D$. Finally, $\star\in D\setminus E$ is a designated
  (non-existing/undefined) object.

  Given a variable assignment $g: \V \longrightarrow D$, we
  define the interpretation function $\|\cdot\|^{I,g}$ for terms and formulas of
  $\FFOL$ as follows:
\noindent
\begin{description}
\item{Terms}
\begin{enumerate}
\item $\| x \|^{I,g} = g(x)$ for variable symbols $x\in\F$
\item $\| c \|^{I,g} = I(c)$, where $c\in\F$ is an 0-ary function symbol
\item $\| f(t_1,\ldots,t_n)\|^{I,g} = I(f)(\| t_1 \|^{I,g},\ldots,\|
  t_n \|^{I,g})$, where $f\in\F$ is an $n$-ary ($n\geq 1$) function symbol
\item $\| \Descr x. r \|^{I,g} = d \in E$, such that $\| r
  \|^{I,g[d/x]} = \True$ and  $\| r
  \|^{I,g[d'/x]} = \False$ for all $d'\not=d\in E$ (i.e. $d$ is the
  unique existing object for which $r$ holds); if there is no
  such $d\in E$, then $\| \Descr x. r \|^{I,g} = \star$
\end{enumerate}

\item{Formulas}
\begin{enumerate}
\item $\| p \|^{I,g} = I(p)$, where $p\in\F$ is an 0-ary predicate
  symbol
\item $\| t_1 = t_2 \|^{I,g} = \True$ iff $\| t_1 \|^{I,g} = \| t_2
  \|^{I,g}$ (this implies that equations such as $1/0 = 1/0$ are
  evaluated to $\True$; however, we have alternative options do define
  this case: e.g. we could
  additionally require that $\| t_1 \|^{I,g} \in E$ and
  $\| t_2 \|^{I,g} \in E$, in which case $1/0 = 1/0$ would be $\False$).
\item $\| p(t_1,\ldots,t_n) \|^{I,g} = \True$ iff $(\| t_1 \|^{I,g},\ldots,\|
  t_n \|^{I,g}) \in I(p)$ for $n$-ary ($n\geq 1$) predicate symbols
  $p\in\P$
\item $\| \neg r \|^{I,g} = \True$ iff $\| r \|^{I,g} = \False$
\item $\| r \Impl s \|^{I,g} = \True$ iff $\| r \|^{I,g} = \False$
  or $\| s \|^{I,g} = \True$
\item $\| \forall x. r  \|^{I,g} = \True$ iff for all $d \in E$ we
  have $\|  r  \|^{I,g[d/x]} = \True$ (here $g[d/x]$ denotes the
  variable assignment that is identical to $g$, except for variable
  $x$, which is now mapped to object $d$)
\end{enumerate}
\end{description}

@@

\subsection{Classical higher-order logic}
\HOL is easily obtained from \HOML by removing the modal operator
$\Box$ from the grammar, and by dropping the set of possible worlds
$W$ and the accessibility relation $R$ from the definition of a model.
Nevertheless, we explicitly state the most relevant definitions for
the particular notion of \HOL as employed in this paper. One reason is
that we do want to carefully distinguish the \HOL and \HOML languages
in the remainder (we use boldface fonts for \HOL and standard fonts
for \HOML). There is also a subtle, but harmless, difference in the
\HOL language as employed here  in comparison to the standard
presentation: here three base types are employed, whereas
usually only two base types are considered. The third base type plays
a crucial role in our embedding of \HOML in \HOL.

\begin{definition}\label{holtypes}
  The set $\hol{T}$ of simple types freely generated from a set of
  basic types $\{\hol{o}, \hol{\mu}, \hol{\iota}\}$ using the function type
  constructor $\hol{\ar}$. $\hol{o}$ is the type of Booleans,
  $\hol{\mu}$ is the type of individuals, and type
  $\hol{\iota}$ is employed as the type of possible worlds below. As
  before we may avoid parentheses.
\end{definition}

\begin{definition}\label{holgrammar}
The grammar for higher-order logic $\HOL$ is:
\begin{align*} 
  \hol{s},\hol{t} \quad ::= \quad & \hol{p_\alpha} \mid \hol{X_\alpha} \mid \hol{(\lam{X_\alpha}
  s_\beta)_{\alpha\ar\beta}} \mid \hol{(s_{\alpha\ar\beta}\, t_\alpha)_\beta} \mid  \hol{\neg_{o\ar o}\, s_o} \mid \\
  & \hol{(({\vee_{o\ar o\ar o}}\,s_o)\,t_o)} \mid \hol{\forall_{(\alpha\ar
    o)\ar o}(\lam{X_\alpha} s_o)} 
\end{align*}
where $\hol{\alpha},\hol{\beta}\in \hol{T}$. The text from
Def.~\ref{homlgrammar} analogously applies, except that we do not
consider the modal  connectives $\hol{\Box}$ and $\hol{\Diamond}$.
\end{definition}

The definitions for substitution (Def.~\ref{homlsubstitution}), $\beta$- and $\eta$-reduction (Def.~\ref{homlbetaeta}), frame (Def.~\ref{homlframe}), and assignment (Def.~\ref{homlassignment}) remain unchanged.

@@

@@

\section{Sound and Complete Embedding in HOL}
Todo ... the embedding in Isabelle/HOL is depicted in 
figure \ref{pic1}

@@

\pagebreak
\newpage

\section{Section title}
\label{sec:1}
Text with citations \cite{RefB} and \cite{RefJ}.
\subsection{Subsection title}
\label{sec:2}
as required. Don't forget to give each section
and subsection a unique label (see Sect.~\ref{sec:1}).
\paragraph{Paragraph headings} Use paragraph headings as needed.
\begin{equation}
a^2+b^2=c^2
\end{equation}

>>>>

%%%%%%%%%%%%%%%%%%%%%%% file template.tex %%%%%%%%%%%%%%%%%%%%%%%%%
%
% This is a general template file for the LaTeX package SVJour3
% for Springer journals.          Springer Heidelberg 2010/09/16
%
% Copy it to a new file with a new name and use it as the basis
% for your article. Delete % signs as needed.
%
% This template includes a few options for different layouts and
% content for various journals. Please consult a previous issue of
% your journal as needed.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% First comes an example EPS file -- just ignore it and
% proceed on the \documentclass line
% your LaTeX will extract the file if required
\begin{filecontents*}{example.eps}
%!PS-Adobe-3.0 EPSF-3.0
%%BoundingBox: 19 19 221 221
%%CreationDate: Mon Sep 29 1997
%%Creator: programmed by hand (JK)
%%EndComments
gsave
newpath
  20 20 moveto
  20 220 lineto
  220 220 lineto
  220 20 lineto
closepath
2 setlinewidth
gsave
  .4 setgray fill
grestore
stroke
grestore
\end{filecontents*}
%
\RequirePackage{fix-cm}
%
%\documentclass{svjour3}                     % onecolumn (standard format)
%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
\documentclass[smallextended]{svjour3}       % onecolumn (second format)
%\documentclass[twocolumn]{svjour3}          % twocolumn
%
\smartqed  % flush right qed marks, e.g. at end of proof
%
\usepackage{amsmath}
\usepackage{xspace,stmaryrd,latexsym,url}
\usepackage{graphicx,tikz,rotating}
\usepackage[strings]{underscore}

%
% \usepackage{mathptmx}      % use Times fonts if available on your TeX system
%
% insert here the call for the packages your document requires
%\usepackage{latexsym}
% etc.
%
% please place your own definitions here and don't use \def but
% \newcommand{}{}
%
% Insert the name of "your journal" with
% \journalname{myjournal}
%

\def\D{\mathit{D}}
\def\E{\mathit{E}}

@@

\newcommand{\imp}{\longrightarrow}
\newcommand{\limp}{\longleftarrow}
\newcommand{\biimp}{\longleftrightarrow}
\newcommand{\allq}{\forall}
\newcommand{\exq}{\exists}
\newcommand{\seq}{\vdash}
\newcommand{\nec}{\Box} % necessarily
\newcommand{\pos}{\Diamond} % possibly
\newcommand{\ess}[2]{#1 \ \mathit{ess.} \ #2}
\newcommand{\NE}{\mathit{NE}}

\def\ti{i}
\def\dom{\textit{dom \/}}
\def\cod{\textit{cod \/}}
\def\comp{\cdot}
\def\kleq{\cong}
\def\exid{\simeq}
\def\ddef{:=}
\def\id{\textit{I\/}}
\def\ex{\E}

@@

\title{Automating Free Logic in HOL, with an Application in Category Theory\thanks{Mention DFG grants ...}}
%\title{Free Logic and Axiom Systems for Category Theory in HOL\thanks{Mention DFG grants ...}}
%\title{Embedding Free Logic and Axiom Systems for Category Theory in HOL\thanks{Mention DFG grants ...}}
%\subtitle{(And an Exemplary Application in Category Theory)}

@@

The purpose of this article is not to claim any novel results in category
theory, but to demonstrate an elegant way to ``implement'' and utilise
reasoning in free logic, and to present respective experiments.

@@

To demonstrate the practical relevance of our new system, we present,
in the second part of this article, a stepwise development of axiom
systems for category theory by generalizing the standard axioms for a
monoid to a partial composition operation. Our purpose is not to make
or claim any contribution to category theory but rather to show how
formalizations involving the kind of logic required (free logic) can
be implemented and validated within modern proof assistants.

A total of eight different axiom systems is studied. The systems I-VI
are shown to be equivalent. The axiom system VII slightly modifies
axiom system VI to obtain (modulo notational transformation) the set
of axioms as proposed by Freyd and Scedrov in their textbook
``Categories, Allegories'' \cite{FreydScedrov90}, published in 1990;
see also Subsection \ref{subsec:FreydNotation} where we present their
original system.  While the axiom systems I-VI are shown to be
consistent, a constricted inconsistency result is obtained for system
VII (when encoded in free logic where free variables range over all
objects): We can prove $\exists x. \neg(\ex x) \Impl False$, where
$\ex$ is the existence predicate. Read this as: If there are undefined
objects, e.g. the value of an undefined composition $x\cdot y$
then we have falsity.  By contraposition, all objects (and thus all
compositions) must exist. But when we assume the latter, then the
axiom system VII essentially reduces categories to monoids.  We note
that axiom system V, which avoids this problem, corresponds to a set
of axioms proposed by Scott \cite{Scott79} in the 1970s. The
problem can also be avoided by restricting the variables in axiom
system VII to range only over existing objects and by postulating
strictness conditions.  This gives us axiom system VIII.

Our exploration has been significantly supported by series of
experiments in which automated reasoning tools have been called from
within the proof assistant Isabelle/HOL via the
Sledgehammer tool. Moreover, we have obtained
very useful feedback at various stages from the model finder Nitpick
\cite{Nitpick} saving us from making several mistakes.

@@

\section{Preliminaries}
\label{sec:preliminaries}
\subsection{Free Logic}
Free logic \cite{Lambert60,Scott67,lambert02:_free_logic} refers to a class of logic
formalisms that are free of basic existence assumptions regarding the
denotation of terms. Remember that terms in e.g. traditional classical
and intuitionistic predicate logics always denote an (existing) object
in a given (non-empty) domain $\D$, and that $\D$ is also exactly the set the
quantifiers range over. In free logic these
basic assumption are abolished. Terms do still denote objects in a
(non-empty) domain $\D$, but a (possibly empty) set $\ex\subseteq \D$ is
chosen to characterize the subdomain of ``existing'' resp. ``defined''
objects in $\D$. Quantification is
now restricted to set $\ex$ of existing/defined objects only. It is
obvious how this can be used to model undefideness and partiality:
problematic terms, e.g. division by zero or improper definite
descriptions, still denote, but they refer to undefined objects, that
is, objects $d$ in $\D\setminus \ex$ lying outside of the scope of
quantification.

The particular notion of free logic as exploited in the remainder of
this article has been proposed by Scott \cite{Scott67}.  A graphical
illustration of this notion of free logic is presented in
Fig.~\ref{fig1}. It
employs a distinguished undefined object $\star$.  
% Free logic (and inclusive logic) has been proposed as an alternative
% to remedy these shortcomings. It distinguishes between a raw domain of
% possibly non-existing objects ‹❙D› and a particular subdomain ‹❙E› of ‹❙D›,
% containing only the ``existing'' entities. Free variables range over ‹❙D›
% and quantified variables only over ‹❙E›. Each term denotes in ‹❙D› but not
% necessarily in ‹❙E›. The particular notion of free logic as exploited below has been
% introduced by Scott @{cite "scott67:_exist"}. A graphical 
% illustration of this notion of free logic is presented in Fig.~\ref{fig1}.
\begin{figure}
\centering
\newcommand\firstellipse{(2,-5) ellipse (6cm and 4cm)}
\newcommand\secondellipse{(0,-5.3) ellipse (3.5cm and 2.5cm)}1
\begin{tikzpicture}
  \begin{scope}[fill opacity=1]
    \fill[gray!40] \firstellipse;
    \fill[gray!10] \secondellipse;
  \end{scope}
  \begin{scope}[very thick,font=\large]
    \draw \firstellipse node[font=\normalsize\bfseries] at (-.5,-5) {\underline{\textbf{$\ex$}: existing objects}};
    \draw \firstellipse node[font=\small] at (-.5,-5.5) {values of bound variables};
    \draw \secondellipse node[font=\normalsize\bfseries] at (4.3,-2.5) {\underline{\textbf{D}: raw objects}};
    \draw \secondellipse node[font=\small] at (4.3,-3) {values of free variables};
  \end{scope}
  \node[font=\normalsize\bfseries] at (6,-5) {$\star$};
  \node[font=\normalsize\bfseries] at (6,-5.3) {undefined};
\end{tikzpicture}
\caption{Illustration of the Semantical Domains of Free Logic \label{fig1}}
\end{figure}

@@

  The \emph{terms and formulas of $\FFOL$} are formally defined as the
  smallest sets such that:
\begin{enumerate}
\item each variable $x\in\V$ is a term of $\FFOL$, 
\item given any $n$-ary ($n\geq 0$) function symbol $f\in\F$ and terms
  $t_1,\ldots,t_n$ of $\FFOL$, then $f(t_1,\ldots,t_n)$ is a term of $\FFOL$, 
\item given terms $t_1$ and $t_2$ of $\FFOL$, then $t_1 = t_2$ is an
  (atomic) formula of $\FFOL$,
\item given any $n$-ary ($n\geq 0$) predicate symbol $p\in\P$ and terms
$t_1,\ldots,t_n$ of $\FFOL$, then $p(t_1,\ldots,t_n)$ is an (atomic)
formula of $\FFOL$, 
\item given formulas $r$ and $s$ of $\FFOL$, then $\neg r$, $r
  \Impl s$ and $\forall x. r$ are (compound) formulas of $\FFOL$
\item given a formula $r$ of $\FFOL$, then $\Descr x. r$ is a
  term of $\FFOL$ (definite  description)
\end{enumerate}
\end{definition}

Further formulas of $\FFOL$, including various defined notions of
equality, can be introduced as abbreviations.

  \emph{Substitution} of a term $r$ for a variable $x$
  in a term $s$ is denoted by $[r/x]s$.

A \emph{variable assignment} $g$ maps
variables $x\in V$ to elements in $D$. $g[d/x]$ denotes the
assignment that is identical to $g$, except for variable $x$, which is
now mapped to $d$.

Regarding the semantics different options have been proposed in the literature. For
example, instead of a possible empty set of existing objects $\ex$, we
could postulate non-emptiness of $\ex$. Here we closely
follow the notion of free logic as proposed by Scott \cite{Scott67}.

\begin{definition}[Dual-domain semantics of $\FFOL$]
  A \emph{model structure for $\FFOL$} consists of a triple
  $\langle D, \ex, I, \star\rangle$, where $D$ is a non-empty raw domain of
  objects, $\ex\subseteq D$ a possible empty set of ``existing''
  resp. ``defined'' objects, and $I$ an interpretation function
  mapping 0-ary function symbols (constants) to defined objects $d\in \ex$, 
 0-ary predicate symbols (propositions) to $\True$ or $\False$, 
  $n$-ary function symbols (for $n\geq 1$) to $n$-ary functions
  $D \times \ldots \times D \longrightarrow D$ and $n$-ary predicate
  symbols (for $n\geq 1$) to $n$-ary relations
  $D \times \ldots \times D$. Finally, $\star\in D\setminus \ex$ is a designated
  (non-existing/undefined) object.

  Given a variable assignment $g$, we
  define the interpretation function $\|\cdot\|^{I,g}$ for terms and formulas of
  $\FFOL$ as follows:
\noindent
\begin{description}
\item{Terms}
\begin{enumerate}
\item $\| x \|^{I,g} = g(x)$ for variable symbols $x\in\V$
\item $\| c \|^{I,g} = I(c)$, where $c\in\F$ is an 0-ary function symbol
\item $\| f(t_1,\ldots,t_n)\|^{I,g} = I(f)(\| t_1 \|^{I,g},\ldots,\|
  t_n \|^{I,g})$, where $f\in\F$ is an $n$-ary ($n\geq 1$) function symbol
\item $\| \Descr x. r \|^{I,g} = d \in \ex$, such that $\| r
  \|^{I,g[d/x]} = \True$ and  $\| r
  \|^{I,g[d'/x]} = \False$ for all $d'\not=d\in \ex$ (i.e. $d$ is the
  unique existing object for which $r$ holds); if there is no
  such $d\in \ex$, then $\| \Descr x. r \|^{I,g} = \star$
\end{enumerate}

\item{Formulas}
\begin{enumerate}
\item $\| p \|^{I,g} = I(p)$, where $p\in\F$ is an 0-ary predicate
  symbol
\item $\| t_1 = t_2 \|^{I,g} = \True$ iff $\| t_1 \|^{I,g} = \| t_2
  \|^{I,g}$ (this basic notion of primitive equality on $D$ implies that equations such as $1/0 = 1/0$ are
  evaluated to $\True$; later, in Section \ref{??}, we will
  define and utilise further notions of equality, including \emph{Kleene equality}
  and \emph{existing equality}).
\item $\| p(t_1,\ldots,t_n) \|^{I,g} = \True$ iff $(\| t_1 \|^{I,g},\ldots,\|
  t_n \|^{I,g}) \in I(p)$ for $n$-ary ($n\geq 1$) predicate symbols
  $p\in\P$
\item $\| \neg r \|^{I,g} = \True$ iff $\| r \|^{I,g} = \False$
\item $\| r \Impl s \|^{I,g} = \True$ iff $\| r \|^{I,g} = \False$
  or $\| s \|^{I,g} = \True$
\item $\| \forall x. r  \|^{I,g} = \True$ iff for all $d \in \ex$ we
  have $\|  r  \|^{I,g[d/x]} = \True$ % (here $g[d/x]$ denotes the
  % variable assignment that is identical to $g$, except for variable
  % $x$, which is now mapped to object $d$)
\end{enumerate}
\end{description}

@@

\subsection{Classical Higher-Order Logic}
Short introduction ... todo.

\begin{definition}\label{holtypes}
  The set $\hol{T}$ of simple types freely generated from a set of
  basic types $\{\hol{o}, \hol{\mu}, \hol{\iota}\}$ using the function type
  constructor $\hol{\ar}$. $\hol{o}$ is the type of Booleans,
  $\hol{\mu}$ is the type of individuals, and type
  $\hol{\iota}$ is employed as the type of possible worlds below. We
  may avoid parentheses if the structure of a complex type is clear in context.
\end{definition}

\begin{definition}\label{holgrammar}
The language of higher-order logic $\HOL$ is defined by the following
grammar:\footnote{Ednote: Maybe better to define this analagous to $\FFOL$.}
\begin{align*} 
  \hol{s},\hol{t} \quad ::= \quad & \hol{p_\alpha} \mid \hol{X_\alpha} \mid \hol{(\lam{X_\alpha}
  s_\beta)_{\alpha\ar\beta}} \mid \hol{(s_{\alpha\ar\beta}\, t_\alpha)_\beta} \mid  \hol{\neg_{o\ar o}\, s_o} \mid \\
  & \hol{(({\vee_{o\ar o\ar o}}\,s_o)\,t_o)} \mid \hol{\forall_{(\alpha\ar
    o)\ar o}(\lam{X_\alpha} s_o)} 
\end{align*}
where $\hol{\alpha},\hol{\beta}\in \hol{T}$. $p_\alpha$ denotes typed constants and
$X_\alpha$ typed variables (distinct from $p_\alpha$).  Complex typed
terms are constructed via abstraction and application. The type of
each term is given as a subscript.  Terms $s_o$ of type $o$ are called
formulas.  The logical connectives of choice are
$\hol{\neg_{o\ar o}}$, $\hol{\vee_{o\ar
  o\ar o}}$, and $\hol{\forall_{(\alpha\ar
  o)\ar o}}$ (for $\hol{\alpha\in T}$).  Type subscripts may be dropped if
irrelevant or obvious. Similarly, parentheses may be avoided.  Binder
notation $\hol{\all{X_\alpha} s_o}$ is used as shorthand for
$\hol{\forall_{(\alpha\ar o)\ar o}
(\lam{X_\alpha} s_o)}$, and infix notation $\hol{s \vee t}$ is employed
instead of $\hol{((\vee s)\, t)}$. From the above connectives, other logical
connectives, such as $\hol{\top}$, $\hol{\bot}$, $\hol{\wedge}$,
$\hol{\supset}$, $\hol{\equiv}$ and $\hol{\exists}$, can be defined in the usual way.
\end{definition}

  \emph{Substitution} of a term $\hol{A_\alpha}$ for a variable $\hol{X_\alpha}$
  in a term $\hol{B_\beta}$ is denoted by $\hol{[A/X]B}$.  Since we consider
  $\hol{\alpha}$-conversion implicitly, we assume the bound variables of $\hol{B}$
  avoid variable capture.

%\begin{definition}\label{homlbetaeta} %\footnote{\textbf{Chris:} Is Def.4 needed at all?}
  Two common relations on terms are given by $\hol{\beta}$-reduction and
  $\hol{\eta}$-reduction.  A $\hol{\beta}$-redex has the form $\hol{(\lam{X}s)t}$ and
  $\hol{\beta}$-reduces to $\hol{[t/X]s}$.  An $\hol{\eta}$-redex has the form
  $\hol{(\lam{X}s X)}$ where variable $\hol{X}$ is not free in $\hol{s}$; it
  $\hol{\eta}$-reduces to $\hol{s}$.  We write $\hol{s\eqb t}$ to mean $\hol{s}$ can be
  converted to $\hol{t}$ by a series of $\hol{\beta}$-reductions and expansions.
  Similarly, $\hol{s\eqbe t}$ means $\hol{s}$ can be converted to $\hol{t}$ using both
  $\hol{\beta}$ and $\hol{\eta}$.  For each $\hol{s_\alpha\in \HOL}$ there is a unique
  \emph{$\beta$-normal form} and a unique \emph{$\beta\eta$-normal
    form}.
% From this fact we know $s\eqb t$ ($s\eqbe t$) iff
% $\Bnormform s\Metaeq\Bnormform t$
% ($\Benormform s\Metaeq\Benormform t$).
% The semantics of $\stt$ is well understood and thoroughly documented
% in the literature \cite{Henkin50,Andrews72b,Andrews72a,BBK04}; our
% summary below is adapted from Andrews~\cite{sep-type-theory-church}.
%\end{definition}

%\begin{definition}\label{homlassignment}
A \emph{variable assignment} $g$ maps
variables $\hol{X_\alpha}$ to elements in $\hol{D_\alpha}$. $\hol{g[d/W]}$ denotes the
assignment that is identical to $\hol{g}$, except for variable $\hol{W}$, which is
now mapped to $\hol{d}$.
%\end{definition}

\begin{definition}\label{homlframe}
A \emph{frame} $\hol{D}$ is a collection $\hol{\{D_\alpha\}_{\alpha\in\entity{T}}}$
of nonempty sets $\hol{D_\alpha}$, such that $\hol{D_o = \{T,F\}}$
(for truth and falsehood).  The
$\hol{D_{\alpha\ar\beta}}$ are collections of functions mapping
$\hol{D_\alpha}$ into $\hol{D_\beta}$. 
\end{definition}

@@

\begin{definition}\label{homlvalid}
A formula $s_o$ is \emph{true} in model $M$ 
under assignment $g$ iff $\|s_o\|^{M,g} = T$; this is also denoted
as $M,g \models^\HOL s_o$.  A formula $s_o$ is called \emph{valid} in
$M$, which is denoted as $M \models^\HOL s_o$, iff $M,g \models^\HOL s_o$ for all
assignments $g$. Finally, a formula $s_o$ is called
\emph{valid}, which we denote by $\models^\HOL s_o$, iff $s_o$ is valid for
all $M$. Moreover, we write $\Gamma\models^\HOL \Delta$ (for sets of
 formulas $\Gamma$ and $\Delta$) iff there is a model $M$ and an assignment $g$ such that  
 $M,g \models^\HOL s_o$ for all $s_o\in \Gamma$ and $M,g \models^\HOL t_o$ for at least one $t_o\in \Delta$.
\end{definition}

@@

\section{Sound and Complete Embedding of $\FFOL$ in HOL}
Todo ... the embedding in Isabelle/HOL is depicted in 
figure \ref{pic1}

@@

In this framework partial and total functions are modelled as follows: 
A function $f$ is total if and only if for all $x$ we have $\ex x \imp \ex(f x)$ 
For partial functions $f$ we may have some $x$ such that $\ex x$ but not
 $\ex(f x)$. A function $f$ is strict  if  and only if for all $x$
we have $\ex(f x) \imp \ex x$.

\section{Exploring Axioms Systems for Category Theory}
In an exemplary theory exploration study, we have shown how Scott's
\cite{Scott79} axiom system for category theory can be derived from a
notion of partial monoids. These axioms systems are presented in Table \ref{axioms-sets-1}.

The stepwise evolution has been described in
\cite{R58}. Below we summarise these experiments.
However, first we describe some basic modeling decisions for the
technical encoding in Isabelle/HOL. The Isabelle/HOL sources of our experiments are
available at \url{todo-sources}.
